A Flexible Machine Learning-Aware Architecture\\ for Future WLANs

### 2020-01-30 LaTeX files/arxiv_submission/main.tex
Update manuscript and README + arXiv files

\usepackage{url}

@@

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Lots of hopes have been placed \textcolor{orange}{on} \textcolor{blue}{machine learning (ML)} as a key enabler of future wireless networks. By taking advantage of the large volumes of data generated by networks, ML is expected to deal with the ever-increasing complexity of networking problems. Unfortunately, current \textcolor{orange}{networks} are not yet prepared \textcolor{orange}{to support} the ensuing requirements of ML-based applications \textcolor{orange}{in terms of} data collection, processing, and output distribution. This article points out the architectural requirements that are needed to pervasively include ML as part of future wireless networks operation. \textcolor{orange}{Specifically, we look into \textcolor{blue}{wireless local area networks (WLANs)}, which, due to their nature, can be found in multiple forms, ranging from cloud-based to edge-computing-like deployments. In particular,} we propose to adopt the International Telecommunications Union (ITU) unified architecture for 5G and beyond. Based on the ITU's architecture, we provide insights on the main requirements and the major challenges of introducing ML to the multiple modalities of WLANs. \textcolor{blue}{Finally, we showcase the superiority of the architecture through an ML-enabled use case for future networks.}  %\textcolor{blue}{Finally, numerical results show the superiority of ML-based methods enabled by the architecture in front of currently adopted approaches.}
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
\textcolor{blue}{Architecture, Future Networks, ITU, Machine Learning, Wireless Local Area Networks}
\end{IEEEkeywords}

@@

\begin{table*}[t!]
	\caption{Machine learning methods, algorithms, potential networking applications, and examples of input data.}
	\label{table:ml_taxonomy}
	\centering
	\begin{tabular}{|p{.09\textwidth}|p{.28\textwidth}|p{.25\textwidth}|p{.25\textwidth}|}
		\hline
		\textbf{ML method} & \textbf{Algorithms} & \textbf{Potential applications} & \textbf{Examples of input data} \\\hline
		Supervised learning & Linear classifiers, regression methods (e.g., \textcolor{blue}{autoregressive integrated moving average}), neural networks, hidden Markov models, random forest, support vector machines, k-nearest neighbors, principal component analysis &Traffic forecasting, mobility pattern prediction, flow classification, routing, anomaly detection, spectrum management, failure detection, QoE prediction & IP traffic matrices, temporal user location, availability of routing paths, application data, channel measurements, performance metrics \\\hline
		Unsupervised learning & Clustering, mixture models, generative models, non-negative matrix factorization, evolutionary algorithms & traffic classification, virtual topology design, path computation, intruder detection, signal separation& IP traffic matrices, historical end-to-end bit-rate, received symbols \\\hline
		Reinforcement learning & Monte Carlo, Q-learning, \textcolor{blue}{state-action-reward-state-action (SARSA)}, deep Q network, actor-critic, multi-armed bandits, learning automaton, \textcolor{blue}{Markov decision processes} & Power control, rate adaptation, routing, channel selection, spatial reuse, smart caching, traffic offloading, cognitive channel access, energy harvesting, \textcolor{blue}{energy efficiency} & Channel measurements, link status, performance metrics (e.g., throughput, delay), \textcolor{blue}{server occupation, power consumption} \\\hline
	\end{tabular}
\end{table*}

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
Wireless communications have reached a point where a paradigm shift is required for satisfying the increasing needs of next-generation applications \cite{osseiran2014scenarios}. Based on the current trend, \textcolor{blue}{artificial intelligence (AI)}, and more precisely \textcolor{blue}{machine learning (ML)}, are expected to conduct a revolution, especially regarding the network planning, operation, and management of the 5th and 6th generation (5G/6G) of mobile communications.

ML is meant to empower a computational system for learning automatically, based on experience, so that future situations can be properly managed without having been programmed explicitly. Concerning wireless communications, there is a huge amount of unexploited data generated at both infrastructure and user levels, which could be extremely useful for learning complex patterns\textcolor{orange}{, which may allow improving network performance}. For instance, the behavior of users in a given network-oriented service can be predicted through ML, based on the information from previous sessions. \textcolor{orange}{Based on these predictions, network resources can be properly accommodated} in future sessions, thus leading to an optimized solution.

@@

The first steps towards AI-enabled networking are currently being made in 5G through network function virtualization (NFV). Unlike traditional hardware-based networks, NFV allows rapid elasticity and fast reconfiguration on assigning network resources, which is particularly useful to enable verticals such as autonomous driving in the automotive sector or smart manufacturing in \textcolor{blue}{industry 4.0}. Besides, network virtualization is useful to boost inter-operator coordination, which would bring the ML operation to a macro-scale level, thus counting with vast information and computation resources.

To conduct the evolution towards ML-aware networks, standardization is key to reach consensus between operators and manufacturers. In this regard, we find many initiatives held by standardization organizations, from which we highlight the Focus Group on Machine Learning for Future Networks including 5G (FG-ML5G), which belongs to the International Telecommunication Union Telecommunication Standardization Sector (ITU-T). The FG-ML5G aims to enable the convergence of future communications with ML technologies. To that end, the focus group has released a specification on a \emph{Unified architecture for 5G and beyond}, which has recently been turned into an ITU Recommendation \cite{itu2019architecture}. The ITU's standardized architecture provides a common nomenclature for ML-related mechanisms so that interoperability with other networking systems is achieved.

Apart from the ITU-T initiatives, other important standardization bodies such as the 3rd Generation Partnership Project (3GPP) or the European Telecommunications Standards Institute (ETSI) are currently working on the integration of data analytics to network functions. The 3GPP contemplates AI as one of the priority topics for shaping its upcoming release (Release 17), and architectural requirements are currently under study \cite{3gpp2019study}. Furthermore, we highlight the ETSI groups on Experiential Networked Intelligence (ENI) and Zero-touch network and Service Management (ZSM), which actively study the integration of AI to networks \cite{etsi2019architecture}. Unlike the ITU's unified architecture, most of the work held by the 3GPP and the ETSI focuses on centralized data collection and data analytics solutions. Nevertheless, we understand that the works in \cite{itu2019architecture, 3gpp2019study, etsi2019architecture} are complementary to each other.
% Alternative architecture: \cite{nekovee2019mobile}

To contribute to the evolution of wireless communications towards AI-based systems, we provide a realization of an ML-aware architecture for IEEE 802.11 \textcolor{blue}{wireless local area networks (WLANs)}, which will be an essential part of the 5G/6G ecosystem in the unlicensed bands. Unlike for cellular networks, WLANs have received much less attention when designing AI-aware architectural solutions. The fact is that cellular-based deployments fit in perfectly with big data analytics, due to the vast amount of data and high computation resources available for mobile network operators (MNOs). In contrast, WLANs pose a set of specific challenges resulting from their multiple deployment modes (e.g., campus network, residential usage) and their typical decentralized nature. Despite WLANs can count with plenty of data to be used by ML methods in large and planned deployments, we find other  \textcolor{orange}{residential-type} scenarios that lack \textcolor{orange}{of} powerful centralized equipment. In  \textcolor{orange}{these} cases, huge computing and processing  \textcolor{orange}{resources} cannot be endowed to the ML operation.

In particular, we adopt the module-based ITU's architecture, which is flexible in terms of deployment heterogeneity and provides adaptability to the problem instance and the set of available resources. This is an important requirement for the integration of different ML-based approaches into the different modalities of WLANs. For instance, despite deep learning is a powerful solution, it entails a set of computation, storage and communication requirements that may not be fulfilled in certain deployments or parts of the network.

@@

\subsection{Machine Learning in Communications}
% 5G and ML as enabler
The proliferation of new communication-based applications is defining the shape of future networks through a set of strict requirements \cite{itu2019use}. Some examples are \textcolor{blue}{vehicle to everything (V2X)}, \textcolor{blue}{industry 4.0}, and \textcolor{blue}{virtual reality / augmented reality (VR/AR)}.  \textcolor{orange}{These} applications are really challenging in terms of bandwidth, latency, reliability, and scalability. In this regard, next-generation wireless networks  \textcolor{orange}{are expected to}:
\begin{enumerate}
	\item Provide a very high data rate (10-20 Gbps).
	\item Support a massive amount of connected devices (1,000,000 devices/km$^2$).
	\item Dramatically reduce the latency ($<5$ ms) whilst maintaining a low packet error rate (1 packet lost for every 10$^5$ packets sent).
\end{enumerate}

In 5G, the previous concepts are referred to as enhanced mobile broadband (eMBB), massive machine to machine and \textcolor{blue}{Internet of things (IoT)} communication (mMTC), and ultra-reliable and low latency communication (uRLLC), respectively. Similarly, IEEE 802.11 groups are also considering those aspects in the design of next-generation amendments, such as high efficiency (HE) IEEE 802.11ax and extreme high throughput (EHT) IEEE 802.11be.

% Overview ML with formal definitions
To meet the abovementioned strict requirements, not only a technological innovation is expected (e.g., use of more spectrum or improve multiple-antennas technologies), but a paradigm shift is necessary when designing solutions for network planning, operation, and management. In particular, intelligent wireless networks need to be empowered with cognitive and context-aware capabilities, which may require additional infrastructure such as environmental sensors and cameras. To that end, ML is expected to be important during the lifetime of 5G and will become pervasive - as included from the beginning in their conception - in 6G networks.

The actual utility of ML lies in those problems that are hard to solve by hand-programming due to their underlying complex patterns (e.g., network traffic prediction). Formally, a machine is said to learn if it improves the performance $\mathbb{P}$ obtained from undertaking task $\mathbb{T}$, based on the gathered experience $\mathbb{E}$ \cite{mitchell1997machine}. Different ML techniques have been categorized in multiple ways, but the most common taxonomy differentiates between supervised learning (labeled data is used for training), unsupervised learning (no labels are used on input data), and reinforcement learning (exploration-exploitation trade-off with label/unlabeled data). Table \ref{table:ml_taxonomy} provides a list of algorithms and potential networking applications for each type of ML \textcolor{blue}{techniques}, as well as some examples of input data  \textcolor{orange}{to be used by these} methods. For further details, we address the interested reader to \cite{jiang2016machine, zhang2019deep, usama2019unsupervised}, which survey a plethora of ML-based applications for networking.

% 1 - Architectural proposals	
Apart from the specific ML solutions to problems in communications, some efforts have been made towards enabling AI-aware networking in more general terms. In this regard, several architectural proposals have been provided so far \cite{bi2015wireless,chih2017big,wang2018machine}. Most of the referenced works agree in the necessary steps for enabling big data analytics in cellular deployments: (1) data collection, (2) data preparation, (3) data analysis, and (4) decision making. Nevertheless, none of these works provide architectural guidelines to introduce ML to wireless networks. In contrast, the ITU's architecture looks deeper into the ML operation and targets the actual procedures involving information gathering, processing, and communication. In addition to the architectural guidelines, the ITU-T has provided a data handling framework for ML-aware networks \cite{itu2019data}, which defines processes concerning data collection, processing, and output distribution.

\subsection{Machine Learning-Enabled Use Cases in WLANs}
To showcase the potential of applying AI in IEEE 802.11 WLANs, we next describe a set of use cases where ML allows improving the network operation and performance.

\subsubsection{OFDMA-Based Smart Network Slicing} 
Network slicing is one of the hottest topics in 5G because it allows separating network resources virtually to meet diverse application requirements. In next-generation WLANs, network slicing can be realized through the allocation of radio resources via orthogonal frequency-division multiple access (OFDMA). However, the heterogeneity of applications running in multiple devices and their subsequent elasticity prevent \textcolor{orange}{allocating} frequency resources easily. \textcolor{orange}{To solve this, ML can be used to make} predictions on the user requirements so that the access network can be optimized.

As an example, Fig. \ref{fig:use_cases} shows a scenario where multiple users \textcolor{orange}{operate under} different requirements, based on the applications they use. While the central controller can make predictions on user behavior, the local schedulers can consider information such as the user profile, the current performance, and the environmental circumstances. In this regard, the \textcolor{blue}{access point (AP)} can allocate the most suitable OFDMA resources to each device, \textcolor{orange}{based on the} predicted needs and \textcolor{orange}{network} status.

\begin{figure}[ht!]
	\centering
	%\hspace{5mm}
	\includegraphics[width=1\columnwidth]{network_slicing_ofdma}
	%\subfigure[Network slicing]{\includegraphics[width=1\columnwidth]{network_slicing_ofdma}\label{fig:network_slicing_ofdma}}
	%	\subfigure[Access Point association]{\includegraphics[width=1\columnwidth]{use_case_ap_association}\label{fig:use_case_ap_association}}
	\caption{OFDMA-based smart network slicing.}
	\label{fig:use_cases}
\end{figure}

\subsubsection{Cloud-Based User Association and Handover}
Most of the current user association and handover procedures held in WLANs typically rely on the strongest signal first (SSF) mechanism. This might be problematic in terms of load balancing \textcolor{orange}{and can potentially lead} to severe performance degradation \textcolor{orange}{in} saturated \textcolor{blue}{basic service sets (BSSs)}. By introducing ML, it is possible to handle contextual information such as the traffic load, which can \textcolor{orange}{be useful for} decision-making. Furthermore, mobility pattern prediction and predicted user requirements can be included in the system, thus empowering the association and handover mechanisms with insightful information. %\textcolor{orange}{Predictions can be provided, for instance, by Deep Neural Networks (DNN).}

%The ML-enabled user association solution is illustrated in Fig. \ref{fig:use_case_ap_association}. A cloud-oriented service obtains data from the different connected WLANs (e.g., number of associated devices, traffic load, performance reports, etc.), which are used for training. Then, new requests are addressed by the ML mechanism, which chooses the best predicted AP for the (re)association, based on the corresponding optimization goals.

\subsubsection{Inference-Based Coordinated Scheduling}
Contrary to traditional cellular-type networks, WLAN deployments can be chaotic, especially in residential scenarios where anyone can set-up an AP and create a wireless network. This typically leads to complex scenarios \textcolor{orange}{where inter-BSS interactions prevent the existing scheduling approaches to ensuring quality of service.} \textcolor{orange}{Fortunately}, ML can be used to infer these interactions and provide a solution accordingly. In particular, through coordinated ML-assisted scheduling, different APs can trigger uplink/downlink transmissions from/to the appropriate stations (STAs), thus increasing the network throughput \textcolor{orange}{whilst} reducing the number of packet collisions.

\subsubsection{Reinforcement Learning-Based Spatial Reuse} 
Spatial reuse aims to improve channel utilization through sensitivity adjustment mechanisms. However, selecting the best sensitivity threshold is not trivial, given the complex spatial interactions that occur among WLANs and the high action space. \textcolor{orange}{To that} purpose, reinforcement learning can be introduced to improve spectral efficiency in a decentralized manner so that each WLAN learns based on experience.

@@

\section{ITU Unified Architecure for Future Networks}
\label{section:itu_architecture}

\begin{figure*}[!ht]
	\centering
	\includegraphics[width=0.5\textwidth]{itu_ml_architecture}
	\caption{ITU logical architecture for future networks \cite{itu2019architecture}. \textcolor{blue}{Entities contain input/output interfaces for communication, while the ML intent is a declarative file with information related to the use case.}}
	\label{fig:itu_ml_architecture}
\end{figure*}

The FG-ML5G was created in November 2017 by its parent group, the ITU-T Study Group 13, \textcolor{orange}{to study} the integration of ML mechanisms into future networks. This includes the definition of interfaces, protocols, data formats, and architectures. During its lifetime, the FG-ML5G has released several reports and contributions. Among them, we highlight the ITU's logical interoperable architecture for future networks \cite{itu2019architecture}, which defines an ML overlay that operates on the top of any unspecified underlay network technology (e.g., 3GPP, EdgeX, IEEE 802.11). The ITU's architecture aims to fulfill a set of technology-agnostic high-level requirements by the utilization of ML. For instance, the architecture must be able to support multiple types of data, thus taking advantage of heterogeneous data sources.

Figure \ref{fig:itu_ml_architecture} shows the elements that compose the ML overlay (management subsystem, multi-level ML pipeline, and closed-loop subsystem), which are further described in the following subsections. Based on this \textcolor{orange}{standard} overlay, ML applications can be instantiated in the logical entities (represented by white boxes).

\subsection{Management Subsystem} 
The management subsystem is in charge of the deployment and the orchestration of the ML services \textcolor{orange}{that operate} in the underlying network. To that purpose, the \textcolor{blue}{machine learning function orchestrator (MLFO)} entity is defined. \textcolor{blue}{The MLFO is first instantiated by a declarative intent that uses a meta language, which specifies the ML use case to be applied, including initialization, policies, and constraints. Then, the MLFO initializes the elements of the multi-level ML pipeline in the corresponding devices and monitors their operation during execution.}

\subsection{Multi-Level Machine Learning Pipeline} 
The multi-level ML pipeline is formed by a set of logical entities that perform the actual ML activity in a given network underlay. The ML pipeline is instantiated and managed by the management subsystem and it is mainly in charge of the data collection and pre-processing, ML model application, and output distribution. The logical entities that compose the ML pipeline are as follows:
\begin{itemize}
	\item \textbf{Source (src):} \textcolor{orange}{generates data to be used by the ML mechanism.}
	\item \textbf{Collector (C):} \textcolor{orange}{collects the data generated by sources.}
	\item \textbf{Pre-processor (PP):} \textcolor{orange}{prepares the data collected for its utilization by the ML mechanism.}
	\item \textbf{Model (M):} \textcolor{orange}{applies an ML model according to the use case.}
	\item \textbf{Policy (P):} \textcolor{orange}{provides a set of constraints and/or guidelines that delimit the behavior of the model.}
	\item \textbf{Distributor (D):} \textcolor{orange}{spreads the ML output across all the corresponding targets (or sinks).}
	\item \textbf{Sink (sink):} \textcolor{orange}{applies the ML output that is received from the distributor.}
\end{itemize}

\subsection{Closed-loop Subsystem} 
In order to address network dynamics, the ML operation is assisted by a closed-loop subsystem. In particular, a sandbox environment can be used to reproduce (internal network) or simulate (simulator/emulator) the effect of certain ML optimization, so that additional information is provided to the system beforehand. As for the ML pipeline, the closed-loop subsystem is orchestrated by the management subsystem. Network simulators (e.g., ns-3, Komondor \cite{barrachina2019komondor}) are examples of closed-loop subsystems, which can serve two purposes: \emph{i)} generate synthetic data to be used for training, and \emph{ii)} \textcolor{blue}{run simulations} for devising the potential of a given ML method to be applied afterward on the real network.

>>>>

\usepackage{hyperref}

@@

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor ns-}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Lots of hopes have been placed on Machine Learning (ML) as a key enabler of future wireless networks. By taking advantage of large volumes of data, ML is expected to deal with the ever-increasing complexity of networking problems. Unfortunately, current networks are not yet prepared to support the ensuing requirements of ML-based applications in terms of data collection, processing, and output distribution. This article points out the architectural requirements that are needed to pervasively include ML as part of future wireless networks operation. Specifically, we look into Wireless Local Area Networks (WLANs), which, due to their nature can be found in multiple forms, ranging from cloud-based to edge-computing-like deployments. In particular, we propose to adopt the International Telecommunications Union (ITU) unified architecture for 5G and beyond. Based on ITU's architecture, we provide insights on the main requirements and the major challenges of introducing ML to the multiple modalities of WLANs. Finally, we showcase the superiority of the architecture through an ML-enabled use case for future networks.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Architecture, Future Networks, ITU, Machine Learning, Wireless Local Area Networks
\end{IEEEkeywords}

@@

\begin{table*}[t!]
	\caption{Machine learning methods, algorithms, potential networking applications, and examples of input data.}
	\label{table:ml_taxonomy}
	\centering
	\begin{tabular}{|p{.09\textwidth}|p{.28\textwidth}|p{.25\textwidth}|p{.25\textwidth}|}
		\hline
		\textbf{ML method} & \textbf{Algorithms} & \textbf{Potential applications} & \textbf{Examples of input data} \\\hline
		Supervised learning & Linear classifiers, regression methods such as Autoregressive Integrated Moving Average (ARIMA), neural networks, hidden Markov models, random forest, support vector machines, k-nearest neighbors, principal component analysis &Traffic forecasting, mobility pattern prediction, flow classification, routing, anomaly detection, spectrum management, failure detection, QoE prediction & IP traffic matrices, temporal user location, availability of routing paths, application data, channel measurements, performance metrics \\\hline
		Unsupervised learning & Clustering, mixture models, generative models, non-negative matrix factorization, evolutionary algorithms & Traffic classification, virtual topology design, path computation, intruder detection, signal separation& IP traffic matrices, historical end-to-end bit-rate, received symbols \\\hline
		Reinforcement learning & Monte Carlo, Q-learning, State-Action-Reward-State-Action (SARSA), deep Q network, actor-critic, multi-armed bandits, learning automaton, Markov decision processes & Power control, rate adaptation, routing, channel selection, spatial reuse, smart caching, traffic offloading, cognitive channel access, energy harvesting, energy efficiency & Channel measurements, link status, performance metrics (e.g., throughput, delay), server occupation, power consumption \\\hline
	\end{tabular}
\end{table*}

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
Wireless communications have reached a point where a paradigm shift is required to satisfy the increasing needs of next-generation applications \cite{osseiran2014scenarios}. Based on the current trend, Artificial Intelligence (AI), and more precisely Machine Learning (ML), is expected to conduct a revolution, especially regarding the network planning, operation, and management of the 5th and 6th generations (5G/6G) of mobile communications.

ML is meant to empower a computational system for learning automatically, based on experience, so that future situations can be properly managed without having been programmed explicitly. Concerning wireless communications, there is a huge amount of unexploited data generated at both infrastructure and user levels, which could be extremely useful for learning complex patterns, thus improving network performance. For instance, the behavior of users in a network-oriented service can be predicted through ML given the information from previous sessions. Based on these predictions, network resources can be appropriately accommodated in future sessions.

@@

The first steps towards AI-enabled networking are currently being made in 5G through Network Function Virtualization (NFV). Unlike traditional hardware-based networks, NFV allows rapid elasticity and fast reconfiguration on assigning network resources. This is particularly useful to enable verticals such as autonomous driving in the automotive sector or smart manufacturing in Industry 4.0. Besides, network virtualization is useful to boost inter-operator coordination and bringing the ML operation to a macro-scale level, counting with vast information and computation resources.

To conduct the evolution towards ML-aware networks, standardization is key to reach consensus between operators and manufacturers. In this regard, we find many initiatives held by standardization organizations, from which we highlight the Focus Group on Machine Learning for Future Networks including 5G (FG-ML5G), which belongs to the International Telecommunication Union Telecommunication Standardization Sector (ITU-T). The FG-ML5G aims to enable the convergence of future communications with ML technologies. To that end, the focus group has released a specification on a \emph{Unified architecture for 5G and beyond}, recently turned into an ITU Recommendation \cite{itu2019architecture}. Remarkably, ITU's standardized architecture provides a common nomenclature for ML-related mechanisms so that interoperability with other networking systems is achieved.

Apart from the ITU-T initiatives, other important standardization bodies such as the 3rd Generation Partnership Project (3GPP) or the European Telecommunications Standards Institute (ETSI) are currently working on the integration of data analytics to network functions. The 3GPP contemplates AI as one of the priority topics for shaping its upcoming release (Release 17) and architectural requirements are currently under study \cite{3gpp2019study}. Furthermore, we highlight the ETSI groups on Experiential Networked Intelligence (ENI) and Zero-touch network and Service Management (ZSM), which actively study the integration of AI to networks \cite{etsi2019architecture}. Unlike the ITU's unified architecture, most of the work held by the 3GPP and the ETSI focuses on centralized data collection and data analytics solutions. Nevertheless, we understand that the works in \cite{itu2019architecture, 3gpp2019study, etsi2019architecture} are complementary to each other.
% Alternative architecture: \cite{nekovee2019mobile}

To contribute to the evolution of wireless communications towards AI-based systems, we provide a realization of the ITU's architecture for IEEE 802.11 Wireless Local Area Networks (WLANs), which will be an essential part of the 5G/6G ecosystem in the unlicensed bands. Unlike for cellular networks, WLANs have received much less attention when designing AI-aware architectural solutions. The fact is that cellular-based deployments fit in perfectly with big data analytics, due to the vast amount of data and high computation resources available for mobile network operators. In contrast, WLANs pose a set of specific challenges resulting from their multiple deployment modes (e.g., campus network, residential usage) and their typical decentralized nature. Despite WLANs can count with plenty of data to be used by ML methods in large and planned deployments, we find other residential-type scenarios that lack of powerful centralized equipment. In these cases, huge computing and processing resources cannot be endowed to the ML operation.

To enable the integration of ML-based approaches into the different modalities of WLANs, the module-based ITU's architecture allows adapting to the problem instance and the set of available resources, thus providing flexibility in terms of deployment heterogeneity. For instance, despite deep learning is a powerful solution that may improve the performance in multiple scenarios, it entails a set of computation, storage and communication requirements that may not be fulfilled in other deployments, or parts of the network.

The main contributions of this paper are as follows:
\begin{itemize}
	\item We devise and discuss the potential of ML-enabled future communications. Then, we focus on IEEE 802.11 WLANs and provide a set of use cases. 
	\item We provide an overview of the ITU's ML-aware architecture for 5G networks and beyond.
	\item We adopt the module-based ITU's architecture and provide a realization for IEEE 802.11 WLANs, thus pointing out the major technical challenges and opportunities.
	\item We depict the potential advantages of ML-based approaches enabled by the architecture through numerical results in a particular use case.	
\end{itemize}

@@

\subsection{Machine Learning in Communications}
% 5G and ML as enabler
The proliferation of new communication-based applications is defining the shape of future networks through a set of strict requirements \cite{itu2019use}. Some examples are Vehicle to Everything (V2X), Industry 4.0, and Virtual Reality / Augmented Reality (VR/AR). These applications are really challenging in terms of bandwidth (10-20 Gbps), latency ($<$5 ms), reliability (1 packet lost for every 10$^5$ packets sent), and scalability (1,000,000 devices/km$^2$), as well as require a flexible network response to cope with the high heterogeneity of devices and contents.

In 5G, the previous concepts are referred to as Enhanced Mobile Broadband (eMBB), Massive Machine to Machine and Internet of Things (IoT) Communication (mMTC), and Ultra-Reliable and Low Latency Communication (uRLLC), respectively. Similarly, IEEE 802.11 groups are also considering these aspects in the design of next-generation amendments, such as High Efficiency (HE) IEEE 802.11ax and Extreme High Throughput (EHT) IEEE 802.11be.

% Overview ML with formal definitions
To meet the abovementioned strict requirements, not only a technological innovation is required (e.g., use of more spectrum or improve multiple-antennas technologies), but a paradigm shift is necessary when designing novel solutions for network planning, operation, and management. In particular, intelligent wireless networks need to be empowered with cognitive and context-aware capabilities, which may require additional infrastructure such as environmental sensors and cameras. To that end, ML is expected to be important during the lifetime of 5G and will become pervasive - as included from the beginning in their conception - in 6G networks.

The actual utility of ML lies in those problems that are hard to solve by hand-programming due to their underlying complex patterns (e.g., network traffic prediction). Formally, a machine is said to learn if it improves the performance $\mathbb{P}$ obtained from undertaking task $\mathbb{T}$, based on the gathered experience $\mathbb{E}$ \cite{mitchell1997machine}. Different ML techniques have been categorized in multiple ways, but the most common taxonomy differentiates between supervised learning (labeled data is used for training), unsupervised learning (no labels are used on input data), and reinforcement learning (exploration-exploitation trade-off with label/unlabeled data). Table \ref{table:ml_taxonomy} provides a list of algorithms and potential networking applications for each type of ML techniques, as well as some examples of input data to be used by these methods. For further details, we address the interested reader to \cite{jiang2016machine, zhang2019deep, usama2019unsupervised}, which survey a plethora of ML-based applications for networking.

% 1 - Architectural proposals	
Apart from the specific ML solutions to problems in communications, some efforts have been made towards enabling AI-aware networking in more general terms. In particular, several architectural proposals have been provided so far \cite{bi2015wireless,chih2017big,wang2018machine}. Most of the referenced works agree in the necessary steps for enabling big data analytics in cellular deployments: (1) data collection, (2) data preparation, (3) data analysis, and (4) decision making. Nevertheless, none of these works provide architectural guidelines to introduce ML to wireless networks. In this regard, the ITU's architecture looks deeper into the ML operation and targets the actual procedures involving information gathering, processing, and communication. Besides, the ITU-T provides a data handling framework for ML-aware networks \cite{itu2019data}, which defines processes concerning data collection, processing, and output distribution.

\subsection{Machine Learning-Enabled Use Cases in WLANs}
To showcase the potential of applying AI in IEEE 802.11 WLANs, we next describe a set of use cases where ML allows improving the network operation.

\subsubsection{OFDMA-Based Smart Network Slicing} 
Network slicing is one of the hottest topics in 5G because it allows virtually separating network resources to meet diverse application requirements. In next-generation WLANs, network slicing can be realized through the allocation of radio resources via Orthogonal Frequency-Division Multiple Access (OFDMA). However, the heterogeneity of applications and devices, and their subsequent elasticity prevent allocating frequency resources easily. To solve this, ML can be used to make predictions on the user requirements so that the access network can be optimized.

As an example, Fig. \ref{fig:use_cases} shows a scenario where multiple users operate under different requirements, based on the applications they use. While the central controller can make predictions on user behavior, the local schedulers may consider information such as the user profile, the current performance, and the environmental circumstances. Accordingly, the Access Point (AP) can allocate the most suitable OFDMA resources to each device, based on the predicted needs and network status.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=1\columnwidth]{network_slicing_ofdma}
	\caption{OFDMA-based smart network slicing.}
	\label{fig:use_cases}
\end{figure}

\subsubsection{Cloud-Based User Association and Handover}
Most of the current user association and handover procedures held in WLANs typically rely on the Strongest Signal First (SSF) mechanism. This might be problematic in terms of load balancing and can potentially lead to severe performance degradation in dense Basic Service Sets (BSSs). By introducing ML, it is possible to handle contextual information such as the traffic load, which can be useful for decision-making. Furthermore, mobility pattern prediction and user requirements forecasts can be included in the system, thus empowering the association and handover mechanisms with insightful information.

\subsubsection{Inference-Based Coordinated Scheduling}
Contrary to traditional cellular-type networks, WLAN deployments can be chaotic, especially in residential scenarios where anyone can set-up an AP and create a wireless network. This typically leads to complex scenarios where inter-BSS interactions prevent the existing scheduling approaches to ensuring a minimum quality of service. Fortunately, ML can be used to infer these interactions and provide a solution accordingly. In particular, through coordinated ML-assisted scheduling, different APs can trigger uplink/downlink transmissions from/to the appropriate stations (STAs), thus increasing the network throughput whilst reducing the number of packet collisions.

\subsubsection{Reinforcement Learning-Based Spatial Reuse} 
Spatial reuse aims to improve channel utilization through sensitivity adjustment mechanisms. However, selecting the best sensitivity threshold is not trivial given the complex spatial interactions that occur among devices. As a potential solution, reinforcement learning can be applied locally to improve spectral efficiency in a decentralized manner.

@@

\section{ITU Unified Architecture for Future Networks}
\label{section:itu_architecture}

\begin{figure*}[!ht]
	\centering
	\includegraphics[width=0.55\textwidth]{itu_ml_architecture}
	\caption{ITU's logical architecture for future networks \cite{itu2019architecture}. Entities contain input/output interfaces for communication, while the ML intent is a declarative file with information related to the use case.}
	\label{fig:itu_ml_architecture}
\end{figure*}

The FG-ML5G was created in November 2017 by its parent group, the ITU-T Study Group 13, to study the integration of ML mechanisms into future networks. This includes the definition of interfaces, protocols, data formats, and architectures. During its lifetime, the FG-ML5G has released several reports and contributions. Among them, we highlight the ITU's logical interoperable architecture for future networks \cite{itu2019architecture}, which defines an ML overlay that operates on the top of any unspecified underlay network technology (e.g., 3GPP, EdgeX, IEEE 802.11). The ITU's architecture aims to fulfill a set of technology-agnostic high-level requirements to support ML. For instance, the architecture must be able to support multiple types of data, thus taking advantage of heterogeneous data sources.

Figure \ref{fig:itu_ml_architecture} shows the elements that compose the ML overlay (management subsystem, multi-level ML pipeline, and closed-loop subsystem). These elements are further described in the following subsections. Based on this standard overlay, ML applications can be instantiated in the logical entities (represented by white boxes).

\subsection{Management Subsystem} 
The management subsystem is in charge of the deployment and the orchestration of the ML services that operate in the underlying network. To that purpose, the Machine Learning Function Orchestrator (MLFO) entity is defined. The MLFO is first instantiated by a declarative intent that uses a meta language. It specifies the ML use case to be applied, including initialization, policies, and constraints. Then, the MLFO initializes the elements of the ML pipeline and monitors their operation during execution.

\subsection{Multi-Level Machine Learning Pipeline} 
The multi-level ML pipeline performs the actual ML operation in a given network underlay and it is in charge of the data collection, model application, and output distribution. The following logical entities compose the ML pipeline:
\begin{itemize}
	\item \textbf{Source (src):} generates data to be used by the ML mechanism.
	\item \textbf{Collector (C):} collects the data generated by sources.
	\item \textbf{Pre-processor (PP):} prepares the data collected for its utilization by the ML mechanism.
	\item \textbf{Model (M):} applies the ML model specified by the intent.
	\item \textbf{Policy (P):} provides a set of constraints and/or guidelines that delimit the behavior of the model.
	\item \textbf{Distributor (D):} spreads the ML output across all the corresponding targets (or sinks).
	\item \textbf{Sink (sink):} applies the ML output that is received from the distributor.
\end{itemize}

\subsection{Closed-loop Subsystem} 
In order to address network dynamics, the ML operation is assisted by a closed-loop subsystem, which can provide information to the system beforehand. As for the ML pipeline, the closed-loop subsystem is orchestrated by the management subsystem. In particular, a sandbox can be formed of real devices (pre-production internal network) or even be virtual (simulator/emulator). Network simulators such as ns-3 and Komondor \cite{barrachina2019komondor} are examples of closed-loop subsystems and can serve two purposes: \emph{i)} generate synthetic data for training, and \emph{ii)} run simulations to validate potential solutions before being applied in production.
